---
title: "Data"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)

theme_set(theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11)))
```

We acquired four datasets for analyzing weather and event impacts on Manhattan subway ridership (2024-2025). Each presented unique technical challenges that required creative solutions.

---

## MTA Subway Ridership

**The Disaggregation Challenge:**

The MTA ridership dataset is highly disaggregated to capture detailed fare collection patterns. Each hourly observation is broken down by:

- **11+ fare class categories**: Full Fare, Senior/Disabled, Students (K-12 and college), Fair Fares (reduced-income riders), and various pass types (7-day, 30-day, unlimited)
- **2 payment methods**: MetroCard (legacy magnetic stripe) and OMNY (contactless tap-to-pay)

This granularity creates a data explosion problem. For Manhattan's 123 stations during a single week:

```
123 stations × 24 hours × 7 days × 11 fare categories × 2 payment methods
= 227,304 records per week
```

For our complete 2024-2025 dataset (22 months ≈ 95 weeks), this would generate **21.6 million records**—far exceeding the Socrata API's 100,000-record pagination limit and creating unwieldy file sizes for analysis.

**The Solution:**

Since our research questions focus on *total ridership* rather than fare type breakdowns, we leveraged the Socrata API's server-side aggregation capabilities. Using `SUM()` with `$group`, we aggregated across all fare types and payment methods *before* data transfer:

```r
"$select" = "transit_timestamp, station_complex_id, station_complex,
             borough, SUM(ridership) as total_ridership",
"$group" = "transit_timestamp, station_complex_id, station_complex, borough"
```

This reduced records by **96%** (from 21.6M to 1.92M), making monthly downloads feasible within API limits while preserving the station-hour granularity needed for our weather and event analyses.

```{r load_ridership, echo=FALSE}
ridership_2024 <- read_csv('data/mta_ridership_full/ridership_2024_manhattan.csv', show_col_types = FALSE)
ridership_2025 <- read_csv('data/mta_ridership_full/ridership_2025_manhattan.csv', show_col_types = FALSE)

ridership <- bind_rows(ridership_2024, ridership_2025) %>%
  mutate(date = as.Date(transit_timestamp), hour = hour(transit_timestamp))
```

This yielded `r format(nrow(ridership), big.mark=",")` hourly observations across `r n_distinct(ridership$station_complex_id)` Manhattan stations from `r min(ridership$date)` to `r max(ridership$date)`.

```{r plot_ridership, fig.height=5, fig.width=10}
# Daily trend (2024 only)
p_daily <- ridership %>%
  filter(year(date) == 2024) %>%
  group_by(date) %>%
  summarize(total = sum(total_ridership, na.rm = TRUE) / 1e6, .groups = "drop") %>%
  ggplot(aes(x = date, y = total)) +
  geom_line(color = "#2E86AB", linewidth = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(title = "Manhattan Subway Daily Ridership (2024)", x = NULL, y = "Daily Ridership (M)") +
  theme(plot.title = element_text(size = 11))

# Hourly pattern
p_hourly <- ridership %>%
  mutate(weekday_type = ifelse(wday(date) %in% 2:6, "Weekday", "Weekend")) %>%
  group_by(hour, weekday_type) %>%
  summarize(avg = mean(total_ridership, na.rm = TRUE) / 1000, .groups = "drop") %>%
  ggplot(aes(x = hour, y = avg, color = weekday_type)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Weekday" = "#2E86AB", "Weekend" = "#E63946")) +
  labs(title = "Average Hourly Pattern", x = "Hour", y = "Avg Ridership (K)", color = NULL) +
  theme(legend.position = "top", plot.title = element_text(size = 11))

# Top stations
p_top <- ridership %>%
  group_by(station_complex) %>%
  summarize(total = sum(total_ridership, na.rm = TRUE) / 1e6, .groups = "drop") %>%
  slice_max(total, n = 30) %>%
  arrange(total) %>%  # Sort by total ascending for proper ordering
  mutate(
    # Truncate station names at first '(' for cleaner display
    station_short = str_replace(station_complex, "\\s*\\(.*", ""),
    # Make names unique by adding row number suffix if duplicates exist
    station_short = make.unique(station_short, sep = " "),
    # Use factor with levels in current order to preserve sorting
    station_short = factor(station_short, levels = station_short)
  ) %>%
  ggplot(aes(x = total, y = station_short)) +
  geom_col(fill = "#06A77D") +
  scale_x_continuous(labels = comma) +
  labs(title = "Top 30 Manhattan Stations", x = "Total Ridership (M)", y = NULL) +
  theme(plot.title = element_text(size = 11), axis.text.y = element_text(size = 8))

p_top | (p_daily / p_hourly)
```

---

## Weather Data

The NOAA API requires an authentication token, which we obtained from [NOAA's token request page](https://www.ncdc.noaa.gov/cdo-web/token). After receiving the token via email, we downloaded daily weather data for Central Park (located in Manhattan) covering 2020-2025. This provides consistent weather measurements applicable to all Manhattan stations.

**Note**: December 2024 data is currently unavailable from the NOAA API, creating a gap in the weather time series. All other months in the 2024-2025 study period have complete daily observations.

```{r load_weather}
weather <- read_csv('data/weather/central_park_weather_wide.csv', show_col_types = FALSE) %>%
  mutate(date = as.Date(date))
```

The dataset contains `r format(nrow(weather), big.mark=",")` daily observations with max temperature, precipitation, and snowfall.

```{r plot_weather, fig.height=7}
# Temperature and precipitation time series
p_ts <- weather %>%
  filter(year(date) >= 2024) %>%
  select(date, TMAX, PRCP) %>%
  pivot_longer(cols = -date, names_to = "metric", values_to = "value") %>%
  mutate(metric = ifelse(metric == "TMAX", "Max Temp (°F)", "Precipitation (in)")) %>%
  ggplot(aes(x = date, y = value)) +
  geom_line(data = . %>% filter(metric == "Max Temp (°F)"), color = "#E63946", linewidth = 0.6) +
  geom_col(data = . %>% filter(metric == "Precipitation (in)"), fill = "#2E86AB", alpha = 0.7) +
  facet_wrap(~metric, ncol = 1, scales = "free_y") +
  labs(title = "Central Park Weather (2024-2025)", x = NULL, y = NULL)

# Seasonal summary
weather_summary <- weather %>%
  filter(year(date) >= 2024) %>%
  mutate(
    season = case_when(
      month(date) %in% c(12, 1, 2) ~ "Winter",
      month(date) %in% c(3, 4, 5) ~ "Spring",
      month(date) %in% c(6, 7, 8) ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
    month_abbr = month(date, label = TRUE, abbr = TRUE)
  )

# Seasonal aggregations
seasonal_weather <- weather_summary %>%
  group_by(season) %>%
  summarize(
    `Avg Temp (°F)` = mean(TMAX, na.rm = TRUE),
    `Storm Days (≥1in)` = sum(PRCP >= 1, na.rm = TRUE),
    `Heat Days (≥90°F)` = sum(TMAX >= 90, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = -season, names_to = "metric", values_to = "value")

# Plot seasonal summary
p_season <- seasonal_weather %>%
  ggplot(aes(x = season, y = value)) +
  geom_col(aes(fill = season)) +
  scale_fill_manual(
    values = c(
      "Winter" = "#457B9D", "Spring" = "#06A77D", "Summer" = "#E63946", "Fall" = "#F77F00"
    )
  ) +
  facet_wrap(~metric, scales = "free", ncol = 3) +
  labs(title = "Seasonal Weather Summary", x = NULL, y = NULL) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))

p_ts / p_season
```

---

## Permitted Events

The NYC Permitted Event Information - Historical dataset (bkfu-528j) contains comprehensive event records but includes hundreds of thousands of small-scale activities (youth sports, film shoots, farmers markets) unlikely to significantly impact subway ridership.

**Smart Filtering Strategy:**

Since the dataset lacks attendance information, we filtered for **large events** (likely 100+ people) using event characteristics as proxies:

- **Event types**: Parades, Athletic Races/Tours, Street Events, Block Parties, Sidewalk Sales
- **Street impact**: Events requiring full street closures
- **Exclusions**: Youth/adult sports leagues, production events, small farmers markets

This filtering reduced from ~2 million total events to **149,524 large events** across NYC.

```{r load_events}
events <- read_csv('data/events/nyc_large_events_2024_2025.csv', show_col_types = FALSE) %>%
  mutate(start_date = as.Date(start_date_time), end_date = as.Date(end_date_time))

events_analysis <- events %>%
  filter(!is.na(start_date)) %>%
  mutate(
    month = month(start_date, label = TRUE),
    year_month = format(start_date, "%Y-%m"),
    borough = case_when(
      str_detect(tolower(event_borough), "manhattan") ~ "Manhattan",
      str_detect(tolower(event_borough), "brooklyn") ~ "Brooklyn",
      str_detect(tolower(event_borough), "queens") ~ "Queens",
      str_detect(tolower(event_borough), "bronx") ~ "Bronx",
      TRUE ~ "Other"
    )
  )
```

Successfully downloaded **`r format(nrow(events), big.mark=",")`** large events (Jan 2024 - Oct 2025), including **`r format(sum(events_analysis$borough == "Manhattan", na.rm=TRUE), big.mark=",")`** in Manhattan.

```{r plot_events, fig.height=5}
# Monthly time series by borough
p_events_time <- events_analysis %>%
  count(year_month, borough) %>%
  mutate(date = as.Date(paste0(year_month, "-01"))) %>%
  ggplot(aes(x = date, y = n, fill = borough)) +
  geom_col() +
  scale_fill_brewer(palette = "Set2") +
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  labs(title = "Large NYC Events by Borough (2024-2025)",
       subtitle = "Strong seasonal pattern: summer/fall peaks, winter lows",
       x = NULL, y = "Number of Events", fill = "Borough") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right")

# Top event types in Manhattan
p_event_types <- events_analysis %>%
  filter(borough == "Manhattan", !is.na(event_type)) %>%
  count(event_type, sort = TRUE) %>%
  slice_max(n, n = 8) %>%
  mutate(event_type = fct_reorder(event_type, n)) %>%
  ggplot(aes(x = n, y = event_type)) +
  geom_col(fill = "#F77F00") +
  scale_x_continuous(labels = comma) +
  labs(title = "Top Event Types in Manhattan", x = "Number of Events", y = NULL) +
  theme(plot.title = element_text(size = 11))

p_events_time / p_event_types
```

---

## Station Locations

Station coordinates are already embedded in the ridership data, so we extracted unique station complexes with coordinates directly from the ridership API rather than seeking a separate geospatial dataset.

```{r load_stations}
stations <- read_csv('data/stations/mta_station_complexes.csv', show_col_types = FALSE)
manhattan_stations <- stations %>%
  filter(borough == "Manhattan", !is.na(latitude), !is.na(longitude))
```

We identified `r nrow(stations)` unique station complexes across NYC, with `r nrow(manhattan_stations)` in Manhattan covering the entire borough from Battery Park to Inwood.

```{r plot_stations, fig.height=7, fig.width=8}
library(leaflet)

# Calculate total ridership by station
station_ridership <- ridership %>%
  group_by(station_complex_id, station_complex) %>%
  summarize(total_ridership = sum(total_ridership, na.rm = TRUE) / 1e6, .groups = "drop")

# Join with station coordinates
manhattan_stations_ridership <- manhattan_stations %>%
  left_join(station_ridership, by = c("station_complex_id", "station_complex")) %>%
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)
  ) %>%
  filter(!is.na(total_ridership))

# Create color palette for ridership
pal <- colorNumeric(
  palette = "YlOrRd",
  domain = log10(manhattan_stations_ridership$total_ridership)
)

# Create interactive map
leaflet(manhattan_stations_ridership) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    lng = ~longitude,
    lat = ~latitude,
    radius = 5,
    color = ~pal(log10(total_ridership)),
    fillColor = ~pal(log10(total_ridership)),
    fillOpacity = 0.8,
    stroke = TRUE,
    weight = 1,
    opacity = 1,
    popup = ~paste0(
      "<strong>", str_replace(station_complex, "\\s*\\(.*", ""), "</strong><br/>",
      "Total Ridership: ", format(round(total_ridership, 2), big.mark = ","), " million"
    )
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~log10(total_ridership),
    title = "Ridership<br/>(Millions, log scale)",
    labFormat = labelFormat(
      transform = function(x) round(10^x, 1)
    )
  ) %>%
  setView(lng = -73.97, lat = 40.76, zoom = 12)
```

---

## Summary

```{r summary_stats, include=FALSE}
total_ridership_millions <- sum(ridership$total_ridership, na.rm=TRUE) / 1e6
heatwave_days <- sum(weather$TMAX >= 90 & year(weather$date) >= 2024, na.rm=TRUE)
storm_days <- sum(weather$PRCP >= 1 & year(weather$date) >= 2024, na.rm=TRUE)
manhattan_events <- sum(events_analysis$borough == "Manhattan", na.rm=TRUE)
```

Our integrated dataset provides comprehensive coverage of Manhattan subway ridership and its potential drivers:

**Ridership:** `r format(nrow(ridership), big.mark=",")` hourly observations across `r n_distinct(ridership$station_complex_id)` Manhattan stations (January 2024–October 2025), totaling over `r round(total_ridership_millions)` million rides. Peak stations like Times Square-42 St and Grand Central-42 St account for a substantial share of daily traffic.

**Weather:** `r format(nrow(weather), big.mark=",")` daily observations from Central Park (2020-2025), including `r heatwave_days` heat wave days (≥90°F) and `r storm_days` major storm events (≥1 inch precipitation) during our study period. Seasonal variation ranges from winter averages in the 30s-40s°F to summer peaks above 90°F.

**Events:** `r format(nrow(events), big.mark=",")` large permitted events (parades, street fairs, athletic races) covering 2024-2025, with `r format(manhattan_events, big.mark=",")` in Manhattan. Events show strong seasonality (summer/fall peaks) and include major NYC happenings (NYC Marathon, street festivals, parades). Smart filtering excluded small-scale events (sports leagues, film shoots) to focus on activities likely to impact subway ridership.

**Stations:** Complete geospatial coverage of `r nrow(manhattan_stations)` Manhattan subway stations, enabling spatial analysis of ridership patterns and proximity-based event impacts.

All four datasets align temporally and spatially, providing comprehensive infrastructure for analyzing **weather and event impacts** on station-level subway ridership across Manhattan's 22-month study period.

**Download scripts:** [`download_full_ridership_monthly.R`](data/download_full_ridership_monthly.R), [`download_historical_events_monthly.R`](data/download_historical_events_monthly.R), [`download_noaa_weather.R`](data/download_noaa_weather.R)
